---
title: "Making a Figure to Illustrate the Genetic Data Permutation Methods"
author: "Eric C. Anderson"
date: "Last Updated: `r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
bibliography: "`r here::here('references.bib')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_knit$set(root.dir = here::here())
start_time <- Sys.time()
```

```{r rmd_snakemake_hacks, include=FALSE}
# here is a little code to make it easy to work interactively with this
# Rmarkdown document to try to find problems.  WHen you run it in snakemake,
# it save the snakemake S4 class object.  When there is no snakemake object,
# it reads it from the file that was previously saved.
rsdir <- ".rmd_snakemake"
dir.create(rsdir, showWarnings = FALSE)
rdsfile <- file.path(rsdir, "admix-fracts.rds")
if(exists("snakemake")) {
  saveRDS(snakemake, file = rdsfile)
} else {
  snakemake <- readRDS(file = rdsfile)
}
```

This is some code to make figures of different ways that one can permute the
genetic data on the founder individuals.

This will also provide a quick check that the methods are doing what we
think they should be.

We will create a simple data set of 12 individuals in three populations 5 in the first,
4 in the second, and 3 in the third.

And we will have markers on three chromosomes.  4 on the first, 5 on the second, and 6 on the third.

We will make that right now, naming alleles according to the haplotypes that
they are on.
```{r}
library(tidyverse)
library(gscramble)

N <- 12
L <- 15
G <- matrix(paste0(1:N, rep(c("a", "b"), each = 12)), nrow = N, ncol = 2 * L)

# make the meta data to go with these
Im <- tibble(
  group = rep(c("pop1", "pop2", "pop3"), times = c(5, 4, 3)),
  indiv = str_c("ind_", 1:N)
)

Mm <- tibble(
  chrom = rep(c("1", "2", "3"), times = c(4, 5, 6)),
) %>%
  group_by(chrom) %>%
  mutate(
    pos = 1500 * 1:n(),
    variant_id = str_c("marker_", chrom, "_", pos)
  )

```

## Preliminaries 

One thing that will be useful is a function to turn a matrix with an even
number of columns into a tibble of alleles in individuals.
```{r}

# turn a matrix with N rows and 2L columns into a tibble
# that names the individual gene copies
gmat2tib <- function(M) {
  if(ncol(M) %% 2 != 0) stop("Must be an even number of columns")
  L <- ncol(M) / 2
  N <- nrow(M)
  
  tibble(
    ind = rep(1:N, times = 2 * L),
    loc = rep(1:L, each = 2 * N),
    hap = rep(rep(c("a", "b"), each = N), times = L),
    ind_hap = factor(str_c(ind, hap), levels = str_c(rep(1:N, each = 2), c("a", "b"))),
    alle = as.vector(M)
  )
  
}

tib <- gmat2tib(G)

chrom_counts <- c(4, 5, 6)
chrom_edges <- c(0, cumsum(chrom_counts))

# now we can plot that.  Let's make a function to do that:
plot_gmat_tib <- function(tib) {
  tib %>% 
    ggplot(aes(x = ind_hap, y = loc)) +
    geom_raster(aes(fill = factor(as.integer(ind)))) +
    geom_tile(fill = NA, color = "black", size = 0.05) +
    geom_text(aes(label = hap)) +
    geom_vline(xintercept = 0.5 + seq(0, 2*N, by = 2)) +
    geom_hline(yintercept = 0.5 + chrom_edges) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) 
}

plot_gmat_tib(tib)
```

But, now, let's permute it using the methods in the `gscramble` package:
```{r}
# this function rearranges the matrix so that
# haplotypes are in columns and markers are in rows
Gr <- rearrange_genos(G, Im, Mm)

# so, we will want to convert that format to a tibble also.
# this function lets us convert ind and hap back to what they were
# originally, after setting ind_hap.  This way we can get the colors
# of the alleles and their original haplotypes that have been permuted into
#the new positions.
rearranged_gmat2tib <- function(M, old_hap_and_ind = FALSE) {
  if(ncol(M) %% 2 != 0) stop("Must be an even number of columns")
  N <- ncol(M) / 2
  L <- nrow(M)
  
  ret <- tibble(
    ind = rep(1:N, each = 2 * L),
    loc = rep(1:L, times = 2 * N),
    hap = rep(rep(c("a", "b"), each = L), times = N),
    ind_hap = factor(str_c(ind, hap), levels = str_c(rep(1:N, each = 2), c("a", "b"))),
    alle = as.vector(M)
  )
  
  if(old_hap_and_ind == TRUE) {
    ret <- ret %>%
      select(-hap, -ind) %>%
      extract(alle, into = c("ind", "hap"), regex = "^([0-9]+)([ab])$")
  }

  ret
}

# then we could read that in
tibR <- rearranged_gmat2tib(Gr$G[[1]])

# and plot it:
plot_gmat_tib(tibR)

```

That all looks just how it should.  

So, now, how about if we permute the markers?  We will do that here, and we will
turn the matrix into a tibble, but after that, we will set ind to the value in
the allele and hap into the a or b from the allele, so the cells are colored
according to what they used to be and the a/b labels are from what they were
before being permuted.
```{r}
set.seed(5)
Gs_plain_permute <- perm_gs_by_pops(Gr)

Gs_plain_permute$G_permed[[1]] %>%
  rearranged_gmat2tib(old_hap_and_ind = TRUE) %>%
  plot_gmat_tib()

```


And, then see what it looks like when we preserve haplotypes:
```{r}
set.seed(5)
Gs_hap_permute <- perm_gs_by_pops(Gr, preserve_haplotypes = TRUE)

Gs_hap_permute$G_permed[[1]] %>%
  rearranged_gmat2tib(old_hap_and_ind = TRUE) %>%
  plot_gmat_tib()

```

OK! That's a wrap!  This is all working as it should, and not I just
have to deal with the colors and then make a good figure out of it.  