---
title: "Preparation for simulating cutthroat-rainbow trout hybrids and passing them to NewHybrids"
author: "Eric C. Anderson"
date: "Last Updated: `r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
bibliography: "`r here::here('references.bib')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_knit$set(root.dir = here::here())
start_time <- Sys.time()
```


Load the libraries.
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(cowplot)


if(!exists("snakemake")) {
  input_list = list(
    summ = "results/compiled/newhybs_sims_pofz.tsv.gz",
    cat = "input/newhybs_categories.txt"
  )
  output_list <- list(
    barplots = "results/summarizing-newhybrids-simulations/newhybs-barplots.pdf",
    barplots_eps = "image-forge/newhybs-barplots.eps"
  )
} else {
  input_list = snakemake@input
  output_list = snakemake@output
}
# you can create the necessary output directories like this:
dump <- lapply(output_list, function(x)
  dir.create(dirname(x), recursive = TRUE, showWarnings = FALSE)
)

```


Let's read in the data and parse the file names, etc.
```{r}
cats <- read_tsv(input_list$cat, col_names = FALSE, skip = 1)
cat_vec <- cats$X1

summ <- read_tsv(input_list$summ, col_names = FALSE)
names(summ) <- c("file", "idx", "id", cat_vec)


# now we pivot it into a long format and parse the file names.  We drop the ones
# that had ids starting with SH or CCT, because those were the learning samples
# in the analysis.  Also, we want to give each individual within nmark values
# a globally unique ID for sorting them...
prepped_summ <- summ %>%
  pivot_longer(-(file:id), names_to = "Z", values_to = "PofZ") %>%
  extract(file, into = c("nmark", "rep"), regex = "results/newhyb_sims/markers_([0-9]+)-rep_([0-9]+)/aa-PofZ.txt", convert = TRUE) %>%
  mutate(true_cat = str_replace(id, "_[0-9]+$", ""), .after = id) %>%
  filter(!(true_cat %in% c("SH", "CCT"))) %>%
  mutate(glob_id = str_c(id, "-", rep))

```

We will probably want to order the true categories in one way or another, so let us do that here in
order of difficulty of inference, sort of,
```{r}
cat_ord <- c(
  "PureCutt",
  "F1",
  "F2",
  "BxC",
  "BxC2",
  "BxCxBxC2",
  "BxC3"
)

```

Now, I want to plot these in pairs---i.e. for each true category I want a 30 markers and a 65 markers
facet together.  And I want to order the individuals in a particular way for each of these pairs:
we are going to group individuals according to the MAP category they are assigned to, and we will
order those groups according to how many individuals are in each, but within each group we will arrange
in descending order on the MAP PofZ.  Cool.
```{r}
# first, get the MAP Z and the MAP PofZ for each fish
map_info <- prepped_summ %>%
  group_by(nmark, rep, id) %>%
  filter(PofZ == max(PofZ)) %>%
  slice(1) %>% # this line in case any were 50/50 (super unlikely!)
  ungroup() %>%
  rename(
    mapZ = Z,
    mapPofZ = PofZ
  )

# now we get the number of individuals in each of those groups and assign them
# a group order.  We do this with the 30 and 65 marker cases combined.
cat_ranks <- map_info %>%
  count(true_cat, mapZ, name = "grp_cnt") %>%
  arrange(true_cat, desc(grp_cnt)) %>%
  group_by(true_cat) %>%
  mutate(grp_rank = 1:n()) %>%
  ungroup()

# now we can join all that back to the prepped_summ and do some arranging and then
# assigning x-values.  
arranged_for_plotting <- prepped_summ %>%
  left_join(
    map_info %>% select(nmark, glob_id, mapZ, mapPofZ), 
    by = join_by(nmark, glob_id)
  ) %>%
  left_join(
    cat_ranks,
    by = join_by(true_cat, mapZ)
  ) %>%
  arrange(nmark, true_cat, grp_rank, desc(mapPofZ)) %>%
  group_by(nmark, true_cat) %>%
  mutate(xpos = as.integer(factor(glob_id, levels = unique(glob_id)))) %>%
  ungroup()
  
```

There is one final thing that we will want to deal with: the colors should be stacked
on the plot so that within each MAP group, the MAP Z color is on the bottom, and then
the other colours will all come in above them.  And when those colors come in above them
we can just leave those in a particular order.  So, let's define some colors and we will
use the names of those to make different factors that describe the order of the colors.

```{r}

# here are some colors:
mycolors <- c(
  CCT = "#a6cee3", 
  BxC2 = "#6a3d9a",
  BxCxBxC2 = "#fb9a99",
  BxC = "#1f78b4",
  F1 = "#fdbf6f", 
  F2 = "#e31a1c",
  BxSHxBxSH2 = "gray80",
  BxSH2 = "gray80",
  BxSH = "gray80", 
  SH = "gray80"
)

Z_ord_basic <- names(mycolors)
```

Now, it turns out that, in these barplots, I want the MAP-Z color to be strictly on the bottom
each time, so that it is easier to see what the total PofZ for that MAP-Z is.  However,
that cannot be implemented with geom_col() because color ordering is done by the
factor levels, and it doesn't seem that you can set those levels for different
groups within the same tibble, and then have them stick correctly when plotting with
ggplot and geom_col().  So, I am just going to "roll my own" bars using geom_rect.

Now, for each individual, we are merely going to order the Z's by the names of mycolors,
after first putting the mapZ one first. And then we will get the endpoints of the rectangles
(stacked as they will be) with a cumsum.
```{r}
further_arranged <- arranged_for_plotting %>%
  mutate(
    Z_f = factor(Z, levels = names(mycolors)),
    NotMapZ = as.integer(!(Z == mapZ))
  ) %>%
  arrange(nmark, true_cat, grp_rank, desc(mapPofZ), NotMapZ, Z_f) %>%
  group_by(nmark, glob_id) %>%
  mutate(ytop = cumsum(PofZ)) %>%
  mutate(ybot = lag(ytop, default = 0), .before = ytop)
  
  
```



Now, we will cycle over the the true categories and make the plots.  

```{r}
plots <- list()
for(TC in cat_ord) {
  tmp <- further_arranged %>%
    filter(true_cat == TC)
  the_plot <- tmp %>%
    ggplot() +
    geom_rect(
      data = tmp,
      mapping = aes(xmin = xpos, xmax = xpos + 1, ymin = ybot, ymax = ytop, fill = Z_f),
      colour = NA,
      linewidth = 0
    ) +
    facet_grid(nmark ~ .) +
    scale_fill_manual(values = mycolors) +
    xlab("Individual Index") +
    ylab("Posterior Prob.") +
    theme_bw() +
    theme(
      legend.position = "none", 
      axis.text.y = element_blank(),
      axis.title.y = element_blank(),
      axis.title.x = element_blank(),
      plot.title = element_text(vjust = -1.2),
      plot.margin = unit(c(0,0,0,0), "pt"),
      panel.spacing = unit(0.5, "mm")
    ) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    ggtitle(TC) 

  plots[[TC]] <- the_plot
    
}
names(plots) <- TCs

# Before we cowplot those together, I am going to have to roll my own legend.  I will
# only include the ones that appear in there.
legie <- tribble( 
  ~cat, ~x, ~y,
  "lCCT",  8, 28,
  "lBxC2", 58, 28, 
  "lF1", 8, 18,
  "lF2", 58, 18,
  "lBxCxBxC2", 8, 8
)
lcol <- mycolors
names(lcol) <- str_c("l", names(mycolors))

legend_bit <- ggplot(legie) +
  geom_rect(aes(xmin = x - 8, xmax = x, ymin = y - 8, ymax = y, fill = cat), colour = "black") +
  geom_text(aes(x = x + 2, y = y - 4, label = cat), hjust = 0, vjust = 0.5) +
  annotate("text", x = 0, y = 32, label = "Inferred Hybrid Category", hjust = 0, vjust = 0.5) +
  annotate("text", x = 45, y = 40, label = "Individual Index", hjust = 0.5, vjust = 0) +
  scale_fill_manual(values = lcol) +
  theme_void() +
  coord_fixed(ratio = 1) +
  guides(fill = guide_legend(position = "none")) +
  theme(legend.position = "none") +
  xlim(-2, 90) +
  ylim(0, 42)
  


# Now, we cowplot it:

plots$LEGEND <- legend_bit
pg <- plot_grid(plotlist = plots, ncol = 1)

ggsave(pg, filename = output_list$barplots, width = 4, height = 10)
ggsave(pg, filename = output_list$barplots_eps, width = 4, height = 10)

system(
  "cd image-forge; latex newyhybs-fig-forge.tex; dvips -E -o newhybs-texed.ps newyhybs-fig-forge.dvi; ps2pdf newhybs-texed.ps; pdfcrop newhybs-texed.pdf"
)
```